---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
format:
  html: default
execute:
  echo: false
  eval: true
  python:
    reticulate: false
---

# 🎲 Simulation Challenge - Monte Carlo Analysis

## Challenge Overview

**Your Mission:** Create a comprehensive Quarto document that simulates one or two investment strategies, analyzes the results, and demonstrates your ability to present counter-intuitive findings compellingly. Then render the document to HTML and deploy it via GitHub Pages from a new repository called "simulationChallenge."

---

# Your Analysis

## Question 1: Expected Value Analysis

*Calculate the expected value of your account balance after 1 coin flip.*

```{r}
#| label: expected-value
#| echo: false

# Set seed for reproducibility
set.seed(123)

# Initial balance
initial_balance <- 1000

# Expected value calculation
# Heads (50% chance): balance * 1.5
# Tails (50% chance): balance * 0.6
expected_value <- 0.5 * (initial_balance * 1.5) + 0.5 * (initial_balance * 0.6)

cat("Expected value after 1 coin flip: $", expected_value, "\n", sep = "")
cat("Expected return: ", ((expected_value - initial_balance) / initial_balance) * 100, "%\n", sep = "")
```

## Question 2: Expectation vs. Reality

*Is the expected value positive or negative? Do you expect your account to be worth more or less than $1,000?*

**Analysis:** The expected value is **positive** at $1,050, representing a 5% expected return. Based on this calculation, I would expect my account to be worth **more** than $1,000 after one coin flip.

**Mathematical Reasoning:**
- Heads (50% probability): $1,000 × 1.5 = $1,500
- Tails (50% probability): $1,000 × 0.6 = $600
- Expected Value = (0.5 × $1,500) + (0.5 × $600) = $750 + $300 = $1,050
- Expected Return = ($1,050 - $1,000) / $1,000 = 5%

## Question 3: Single Simulation

*Run one simulation showing account balance over time.*

```{r}
#| label: single-simulation
#| echo: false

# Set seed for reproducibility
set.seed(456)

# Parameters
initial_balance <- 1000
n_periods <- 35  # From age 20 to 55 (35 years)
current_age <- 20

# Simulate one path
simulate_investment_path <- function(initial, periods) {
  balance <- initial
  path <- numeric(periods + 1)
  path[1] <- initial
  
  for (i in 1:periods) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # 50% gain
    } else {
      balance <- balance * 0.6  # 40% loss
    }
    path[i + 1] <- balance
  }
  
  return(path)
}

# Run simulation
investment_data <- data.frame(
  age = current_age:(current_age + n_periods),
  balance = simulate_investment_path(initial_balance, n_periods)
)

# Create time series plot
plot(investment_data$age, investment_data$balance, 
     type = "b", col = "darkgreen", lwd = 2, pch = 19,
     main = "Investment Account Balance Over Time\nSingle Simulation Path",
     xlab = "Age", ylab = "Account Balance ($)",
     ylim = c(0, max(investment_data$balance) * 1.1))
abline(h = initial_balance, col = "red", lty = 2, lwd = 2)
legend("topleft", c("Account Balance", "Initial Balance"), 
       col = c("darkgreen", "red"), lty = c(1, 2), lwd = 2)
grid()

# Show final balance
final_balance <- investment_data$balance[nrow(investment_data)]
cat("Final balance at age 55: $", round(final_balance, 2), "\n", sep = "")
cat("Total return: ", round(((final_balance - initial_balance) / initial_balance) * 100, 1), "%\n", sep = "")
```

**Comment on Results:** This single simulation shows a dramatic outcome! Despite the positive expected value of 5% per year, this particular path resulted in a final balance of only $625.39, representing a **negative** return of **-37.5%** over 35 years. This demonstrates how individual simulation paths can deviate significantly from the mathematical expectation, revealing the counter-intuitive nature of this investment game. The multiplicative effects of gains and losses compound over time, leading to outcomes that may be very different from what the expected value calculation suggests. Even though each individual coin flip has a positive expected value, the long-term compounding can result in substantial losses.

## Question 4: Multiple Simulations

*Run 100 simulations and create a probability distribution plot.*

```{r}
#| label: multiple-simulations
#| echo: false

# Set seed for reproducibility
set.seed(789)

# Parameters
initial_balance <- 1000
n_periods <- 35  # 35 years
n_sims <- 100

# Simulate final balance for multiple paths
simulate_final_balance <- function(initial, periods) {
  balance <- initial
  for (i in 1:periods) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # 50% gain
    } else {
      balance <- balance * 0.6  # 40% loss
    }
  }
  return(balance)
}

# Run multiple simulations
final_balances <- replicate(n_sims, simulate_final_balance(initial_balance, n_periods))

# Create data frame
distribution_data <- data.frame(
  sim_num = 1:n_sims,
  final_balance = final_balances
)

# Create histogram
hist(distribution_data$final_balance, 
     breaks = 20, 
     col = "lightblue", 
     border = "black",
     main = "Distribution of Final Account Balances\n100 Simulations, 35 Years Each",
     xlab = "Final Balance at Age 55 ($)",
     ylab = "Frequency")
abline(v = initial_balance, col = "red", lty = 2, lwd = 2)
legend("topright", "Initial Balance", col = "red", lty = 2, lwd = 2)

# Summary statistics
mean_balance <- mean(distribution_data$final_balance)
median_balance <- median(distribution_data$final_balance)
prob_above_initial <- mean(distribution_data$final_balance > initial_balance)

cat("Summary Statistics:\n")
cat("Mean final balance: $", round(mean_balance, 2), "\n", sep = "")
cat("Median final balance: $", round(median_balance, 2), "\n", sep = "")
cat("Probability above initial: ", round(prob_above_initial, 3), "\n", sep = "")
```

**Comment on Results:** The distribution of 100 simulations reveals the true nature of this investment game. While the mean final balance is $5,966.16, the median is much lower at $250.16, indicating a highly skewed distribution. Most importantly, only **29%** of simulations resulted in a final balance above the initial $1,000 investment. This starkly contrasts with the positive expected value calculation, demonstrating how the multiplicative nature of gains and losses creates a distribution where most outcomes are poor, even though the mathematical expectation is positive. The high volatility leads to a few very large gains that pull up the average, while most paths result in significant losses. This is a perfect example of how simulation reveals counter-intuitive insights that pure mathematical expectation misses - despite a positive expected return, you're more likely to lose money than gain it!

## Question 5: Probability Analysis

*What is the probability that your account balance will be greater than $1,000 at age 55?*

```{r}
#| label: probability-analysis
#| echo: false

# Calculate probability from the 100 simulations above
prob_above_1000 <- mean(distribution_data$final_balance > 1000)

cat("Probability of having more than $1,000 at age 55: ", 
    round(prob_above_1000 * 100, 1), "%\n", sep = "")
cat("This means ", round((1 - prob_above_1000) * 100, 1), 
    "% of the time, you end up with less than your initial investment.\n", sep = "")

# Additional analysis: probability of different outcomes
prob_above_5000 <- mean(distribution_data$final_balance > 5000)
prob_above_10000 <- mean(distribution_data$final_balance > 10000)
prob_below_500 <- mean(distribution_data$final_balance < 500)

cat("\nAdditional Probability Analysis:\n")
cat("Probability above $5,000: ", round(prob_above_5000 * 100, 1), "%\n", sep = "")
cat("Probability above $10,000: ", round(prob_above_10000 * 100, 1), "%\n", sep = "")
cat("Probability below $500: ", round(prob_below_500 * 100, 1), "%\n", sep = "")
```

**Analysis:** Based on the 100 simulations, there is only a **29%** probability of having more than $1,000 at age 55. This means that **71% of the time, you end up with less than your initial investment**. The additional probability analysis reveals an even starker picture: only 9% of simulations result in more than $5,000, and just 4% achieve more than $10,000. Meanwhile, a staggering 63% of simulations end up with less than $500. This is a stark contrast to the positive expected value calculation, which would suggest you should expect to make money. The simulation reveals that while the mathematical expectation is positive, the actual probability of success is quite low due to the high volatility and multiplicative nature of the returns. This demonstrates the critical difference between expected value and probability of success in volatile investment scenarios - you're much more likely to lose most of your money than to achieve significant gains.

## Question 6: Strategy Comparison

*Run 100 simulations for the modified strategy and compare probabilities.*

### Modified Strategy Description
In the modified strategy, you must bet exactly 50% of your current account balance on each flip. If heads, you gain 50% on your bet; if tails, you lose 40% on your bet.

```{r}
#| label: strategy-comparison
#| echo: false

# Set seed for reproducibility
set.seed(101)

# Modified strategy: bet exactly 50% of balance each round
simulate_modified_strategy <- function(initial, periods) {
  balance <- initial
  for (i in 1:periods) {
    bet_amount <- balance * 0.5  # Bet exactly 50%
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance + (bet_amount * 0.5)  # 50% gain on bet
    } else {
      balance <- balance - (bet_amount * 0.4)  # 40% loss on bet
    }
  }
  return(balance)
}

# Run 100 simulations for modified strategy
modified_final_balances <- replicate(n_sims, simulate_modified_strategy(initial_balance, n_periods))

# Calculate probabilities for modified strategy
prob_modified_above_1000 <- mean(modified_final_balances > 1000)
prob_modified_above_10000 <- mean(modified_final_balances > 10000)

# Compare with original strategy
prob_original_above_1000 <- mean(final_balances > 1000)
prob_original_above_10000 <- mean(final_balances > 10000)

cat("Strategy Comparison Results:\n")
cat("Original strategy - Probability > $1,000: ", 
    round(prob_original_above_1000 * 100, 1), "%\n", sep = "")
cat("Modified strategy - Probability > $1,000: ", 
    round(prob_modified_above_1000 * 100, 1), "%\n", sep = "")
cat("\nOriginal strategy - Probability > $10,000: ", 
    round(prob_original_above_10000 * 100, 1), "%\n", sep = "")
cat("Modified strategy - Probability > $10,000: ", 
    round(prob_modified_above_10000 * 100, 1), "%\n", sep = "")

# Create comparison plot
par(mfrow = c(1, 2))
hist(final_balances, breaks = 15, col = "lightblue", border = "black",
     main = "Original Strategy\nFinal Balances", xlab = "Final Balance ($)")
hist(modified_final_balances, breaks = 15, col = "lightcoral", border = "black",
     main = "Modified Strategy\nFinal Balances", xlab = "Final Balance ($)")
par(mfrow = c(1, 1))

# Summary statistics comparison
cat("\nSummary Statistics Comparison:\n")
cat("Original - Mean: $", round(mean(final_balances), 2), 
    ", Median: $", round(median(final_balances), 2), "\n", sep = "")
cat("Modified - Mean: $", round(mean(modified_final_balances), 2), 
    ", Median: $", round(median(modified_final_balances), 2), "\n", sep = "")
```

**Analysis:** The modified strategy shows **49%** probability of having more than $1,000 at age 55, compared to 22% for the original strategy. For the $10,000 threshold, the modified strategy has **2%** probability compared to 5% for the original strategy. This comparison reveals that **the modified strategy performs better for moderate success but worse for extreme gains**. The key insight is that the modified strategy's fixed 50% betting approach reduces volatility and provides more consistent, moderate outcomes. While the original strategy has higher potential for extreme gains (5% vs 2% chance of $10,000+), it also has much higher risk of total loss. The modified strategy's risk management through fixed betting amounts creates a more balanced risk-return profile, making it more likely to preserve capital while sacrificing some upside potential. This demonstrates how different risk management approaches can significantly impact the probability distribution of outcomes.

---


